### 2021-12-11 thesis/tex/oscillations.tex
update

\begin{figure}[!h]
    \begin{center}
    \includegraphics[width=1\textwidth]{figures/vac_osc.pdf}
    \caption{Probabilities of $\nm$ oscillations after travelling 12 000 km in vaccum.}\label{fig:vac_osc}
    \end{center}
\end{figure}

@@

where the projection operator $(1- \gamma^5)$ in the first line ensures that only the left-hand compontent of the neutrino fields interact.

@@

Using a value of $Y=0.5$ electron per nucleon, we express the matter potential as
\begin{align}
    V_{CC} &= \sqrt{2}G_F N_e = \sqrt{2}G_F Y N_A \,\rho \nonumber \\
           &= \SI{3.8e-23}{\eV} \times \rho\,,
\end{align}
a low number due to the smallness of $G_F$. This is plotted in the right panel of Fig.~\ref{fig:potential}
\begin{figure}
    %TODO: labels
    \centering
    \includegraphics[width=0.9\textwidth]{figures/potential.pdf}
    \caption{\emph{Left panel:} The spherically symmetric Earth density according to PREM~\cite{PREM}, as of distance from the core.
    \emph{Right panel:} $V_{CC}$ using the PREM density and 1/2 electrons per nucleon.}\label{fig:potential}
\end{figure}

@@

\begin{figure}
    %TODO: labels
    % Compare with vacuum oscillation plot and a two neutrino plot. Then we see for certain E we see matter effects more and for some less.
    \centering
    \includegraphics[width=0.7\textwidth]{figures/oscillations.pdf}
    \caption{Neutrino oscillation probabilities after traversing the diameter of the Earth. 
    Oscillation parameters are taken from NuFit~\cite{nufit} and given in Eq.~\ref{eq:PINGUparams}.}\label{fig:oscillations}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/oscillograms.pdf}
    \caption{Oscillograms showing neutrino oscillations for all flavors in $E - \cos(\theta_z)$ space. Note the panels are symmetric about the diagonal panels, i.e. $P_{\a\b} = P_{\b\a}$ due to $\delta_{CP} = 0$. The core-mantle boundary from Fig.~\ref{fig:potential} is clearly displayed at $\cos{(\theta_z)} = -0.83$ as a sharp discontinuity for all flavors.}\label{fig:oscillograms}%TODO: redo finer
\end{figure}

>>>>

\begin{figure}[!h]
    \begin{center}
    \includegraphics[width=1\textwidth]{figures/vac_osc.pdf}
    \caption{Probabilities of $\nm$ oscillations after travelling 12 000 km in vacuum.}\label{fig:vac_osc}
    \end{center}
\end{figure}

@@

where the projection operator $(1- \gamma^5)$ in the first line ensures that only the left-hand component of the neutrino fields interact.

@@

Using a value of $Y=0.5$ electron per nucleon, we express the matter potential as
\begin{align}
    V_{CC} &= \sqrt{2}G_F N_e = \sqrt{2}G_F Y N_A \,\rho \nonumber \\
           &= \SI{3.8e-23}{\eV} \times \rho\,,
\end{align}
a low number due to the smallness of $G_F$. This is plotted in the right panel of Fig.~\ref{fig:potential}
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/potential.pdf}
    \caption{\emph{Left panel:} The spherically symmetric Earth density according to PREM~\cite{PREM}, as of distance from the core.
    \emph{Right panel:} $V_{CC}$ using the PREM density and 1/2 electrons per nucleon.}\label{fig:potential}
\end{figure}

@@

\begin{figure}
    % Compare with vacuum oscillation plot and a two neutrino plot. Then we see for certain E we see matter effects more and for some less.
    \centering
    \includegraphics[width=0.9\textwidth]{figures/oscillations.pdf}
    \caption{Neutrino oscillation probabilities after traversing the diameter of the Earth. 
    Oscillation parameters are taken from NuFit~\cite{nufit} and given in Eq.~\ref{eq:PINGUparams}.}\label{fig:oscillations}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/oscillograms.pdf}
    \caption{Oscillograms showing neutrino oscillations for all flavors in $E - \cos(\theta_z)$ space using parameters from~\cite{nufit}. Note the panels are symmetric about the diagonal panels, i.e. $P_{\a\b} = P_{\b\a}$ due to $\delta_{CP} = 0^\circ$. The core-mantle boundary from Fig.~\ref{fig:potential} is clearly displayed at $\cos{(\theta_z)} = -0.83$ as a sharp discontinuity for all flavors.}\label{fig:oscillograms}%TODO: redo finer
\end{figure}

### 2021-12-11 thesis/tex/sterileResults.tex
update

% \documentclass{thesis}
% \usepackage[margin=0.5in]{geometry}
% \input{commands}

% \begin{document}
\section{The Sterile Hypothesis}\label{sec:sterileResults}
With the null ($3\nu$) hypothesis normalized to the IceCube Monte Carlo as in Eq.~\ref{eq:Nth}, we are now in good shape to study the sterile effect on the probabilities, and how that compares to data.
First, let us see how the collected data from Fig.~\ref{fig:IC_data} deviates from the predicted $3\nu$ oscillations.
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/IC_rates.pdf}
    \caption{How the null hypothesis compares to the IceCube data from~\cite{IC2020}.}\label{fig:IC_rates}
\end{figure}
Fig.~\ref{fig:IC_rates} shows an impressive agreement to the standard $3\nu$ oscillation picture, with the largest deviations of $~6\%$
for neutrinos with high single-digit \si{\TeV} values. In zenith, the data only deviates $\pm 2\%$. Recalling the \si{\TeV} disappearance from 
Fig.~\ref{fig:sterile_resonance}, we see that we have a similar deficiency found in the data at \SI{4}{\TeV}. 
%Thus, we expect the best fitting 
%sterile hypothesis to include a $\dm[41]$ that places the resonant disappearance in the same region. From Fig.~\ref{fig:resonance_shift}, we see
%that a mass of $\dm[41] = \SI{1.5}{\eV \squared}$ achieves this.

\subsection{\texorpdfstring{$\chi^2$}{Chi-squared} Minimization}
For our analyses, we define our $\chi^2$ as
\begin{align} \label{eq:chisq}
    \chi^{2}(\hat{\theta},\alpha,\beta)=\sum_{ij} \frac{\left(N^\text{th}-N^\text{data}\right)_{ij}^{2}}
    {\left(\sigma^\text{data}_{ij}\right)^{2} + \left(\sigma^\text{syst}_{ij}\right)^{2}}+ 
    \frac{(1-\alpha)^2}{\sigma_\alpha^2} + \frac{\beta^2}{\sigma_\beta^2}\,
\end{align}
where we minimize over the model parameters $\hat{\theta}$, the penalty terms $\alpha$ and $\beta$.
$N_{ij}^\text{th}$ is the expected number of events from theory, and $N_{ij}^\text{data}$ is the observed number of events in that bin. 
We set $\sigma_\alpha = 0.25$ as the atmospheric flux normalization error, and $\sigma_\beta = 0.05$ as the zenith angle slope error~\cite{hondapaper}. 
The observed event number has an associated Poissonian uncertainty $\sigma_{ij}^\text{data} = \sqrt{N_{ij}^\text{data}}$.
For IceCube, the event count takes the form
\begin{align}
    N^\text{th}_{ij} = \alpha\left[1+\beta (0.5 + \zreco_i )\right] N_{ij}(\hat{\theta})\,,
\end{align}
with $N_{ij}(\hat{\theta})$ from Eq.~\ref{eq:Nth}. Here, the term $ \beta (0.5 + \zreco_i )$ allows the event distribution to rotate with angle $\beta$ around the median zenith angle of $\zreco = -0.5$.
We also have an uncorrelated systematic error $\sigma_{ijk}^\text{syst}$. We set $\sigma_{ijk}^\text{syst} = f\sqrt{N_{ijk}^\text{data}}$, where $f$ is a factor of our own choosing.

@@

\subsection{Sterile Mass and Mixing}
Let us start with looking at the best-fit event distribution resulting from the $\chi^2$ minimization.
Fig.~\ref{fig:final_rate_plot} now contains the best-fit event distribution assuming the sterile hypothesis.
The best-fit values are $\dm[41] = \SI{0.01}{\eV\squared}$ and 
$\sin(2\theta_{24})^2 = 0.67$ ($\theta_{24} = \SI{27.5}{\degree}$). %TODO:double check these
The contour plot shown in Fig.~\ref{fig:s24_contour} divides the parameter space into two regions.
To the right of the boundary, the $\Delta \chi^2$ has values above the confidence level, meaning that 
those parameter pairs can be excluded at a certain confidence level.

By tuning the scaling $f$ of the uncorrelated systematic error, we can shift the contour. We extract the contour from the 
IceCube sterile analysis~\cite{IC2020}, and compare in Fig.~\ref{fig:error_tuning}. We see that the contour in the 
higher $\dm[41]$ region above \SI{1}{\eV\squared} is very sensetive to the tuning, while the lower region between \SIrange[]{1e-1}{1}{\eV\squared}
remains fixed. %We put $f=35\%$ and compare our results to the Giunti et al.~\cite{gariazzoUpdatedGlobalAnalysis2017} in Fig.~\ref{fig:s24_contour}. %TODO fix
%Here, we see that our lower fixed region between \SIrange{1e-1}{1}{\eV\squared} is closer to the IceCube result than Giunti et al, but that 
%our contour slightly diverges above \SI{1}{\eV\squared}. 
\begin{figure}
    \centering
    \includegraphics[scale=0.58]{figures/s24_error_tuning.pdf}
    \caption{The 90\% CL contour with two different systematic error scalings, compared 
    with the official IceCube result from~\cite{IC2020}.}\label{fig:error_tuning}
\end{figure}

%\begin{figure}
%    \centering
%    \includegraphics[width=0.8\textwidth]{figures/s24_contour.pdf}
%    \caption{The 90\% CL contour using $f=35\%$, compared with the official IceCube result from~\cite{IC2020}
%    %and the result from~\cite{gariazzoUpdatedGlobalAnalysis2017}
%    }\label{fig:s24_contour}    
%\end{figure} %TODO fix

We now in Fig.~\ref{fig:final_rate_plot} plot the best-fit sterile hypothesis with the data from Fig.~\ref{fig:IC_rates}, to finally see how the best 
sterile hypothesis compares to data.
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/final_rate_plot.pdf}
    \caption{The predicted binned event count assuming our best-fit sterile hypothesis, compared with the 
    null and data.}\label{fig:final_rate_plot}
\end{figure}

%\end{document}

>>>>

\section{The Sterile Hypothesis}\label{sec:sterileResults}
With the null ($3\nu$) hypothesis normalized to the IceCube Monte Carlo as in Eq.~\ref{eq:MC_norm}, we are now in good shape to study the sterile neutrino effect on the probabilities, and how that compares to data
\footnote{We do not consider DeepCore nor PINGU in the sterile neutrino analysis because the $\Pamam$ resonance does not appear in the \si{\GeV} region for \si{\eV^2}-scale $\dm[41]$, as shown in Fig.~\ref{fig:sterile_resonance}.
We bring DeepCore and PINGU back in the next section.}.
First, let us see how the collected data from Fig.~\ref{fig:IC_data} deviates from the predicted $3\nu$ oscillations.
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/IC_rates.pdf}
    \caption{How the IceCube null hypothesis from~\cite{IC2020} compares to their collected data.}\label{fig:IC_rates}
\end{figure}
Fig.~\ref{fig:IC_rates} shows an impressive agreement to the standard $3\nu$ oscillation picture, with the largest deviations of $~6\%$
for neutrinos with high single-digit \si{\TeV} values. In zenith, the data only deviates $\pm 2\%$.

\subsection{\texorpdfstring{$\chi^2$}{Chi-squared} Minimization}
For our analyses, we define our $\chi^2$ as
\begin{align} \label{eq:chisq}
    \chi^{2}(\hat{\theta},\alpha,\beta)=\sum_{ij} \frac{\left(N^\text{th}-N^\text{data}\right)_{ij}^{2}}
    {\left(\sigma^\text{data}_{ij}\right)^{2} + \left(\sigma^\text{syst}_{ij}\right)^{2}}+ 
    \frac{(1-\alpha)^2}{\sigma_\alpha^2} + \frac{\beta^2}{\sigma_\beta^2}\,
\end{align}
where we minimize over the model parameters $\hat{\theta}$, the penalty terms $\alpha$ and $\beta$.
$N_{ij}^\text{th}$ is the expected number of events from theory, and $N_{ij}^\text{data}$ is the observed number of events in that bin. 
We set $\sigma_\alpha = 0.25$ as the atmospheric flux normalization error, and $\sigma_\beta = 0.05$ as the zenith angle slope error~\cite{hondapaper}. 
The observed event number has an associated Poissonian uncertainty $\sigma_{ij}^\text{data} = \sqrt{N_{ij}^\text{data}}$.
For IceCube, the event count takes the form
\begin{align}
    N^\text{th}_{ij} = \alpha\left[1+\beta (0.5 + \zreco_i )\right] N_{ij}(\hat{\theta})\,,
\end{align}
with $N_{ij}(\hat{\theta})$ from Eq.~\ref{eq:MC_norm}. Here, the term $ \beta (0.5 + \zreco_i )$ allows the event distribution to rotate with angle $\beta$ around the median zenith angle of $\zreco = -0.5$.
We also have an uncorrelated systematic error $\sigma_{ijk}^\text{syst}$. We set $\sigma_{ijk}^\text{syst} = f\sqrt{N_{ijk}^\text{data}}$, where $f$ is a factor of our own choosing.

@@

\subsection{Sterile Mass and Mixing}
Let us start with looking at the best-fit event distribution resulting from the $\chi^2$ minimization.
Fig.~\ref{fig:final_rate_plot} now contains the best-fit event distribution assuming the 3+1 hypothesis.
The best-fit values are $\dm[41] = \SI{0.01}{\eV\squared}$ and 
$\sin(2\theta_{24})^2 = 0.67$ ($\theta_{24} = \SI{27.5}{\degree}$). %TODO: double check these
The contour plot shown in Fig.~\ref{fig:error_tuning} divides the parameter space into two regions.
To the right of the boundary, the $\Delta \chi^2$ has values above the confidence level, meaning that 
those parameter pairs can be excluded at a certain confidence level.

By tuning the scaling $f$ of the uncorrelated systematic error, we can shift the contour. We extract the contour from the 
IceCube sterile neutrino analysis~\cite{IC2020}, and compare in Fig.~\ref{fig:error_tuning}. We see that the contour in the 
higher $\dm[41]$ region above \SI{1}{\eV\squared} is very sensitive to the tuning, while the lower region between \SIrange[]{1e-1}{1}{\eV\squared}
remains fixed.
\begin{figure}
    \centering
    \includegraphics[scale=0.58]{figures/s24_error_tuning.pdf}
    \caption{The 90\% CL contour with two different systematic error scalings, compared 
    with the official IceCube result from~\cite{IC2020}.}\label{fig:error_tuning}
\end{figure}

We now in Fig.~\ref{fig:final_rate_plot} plot the best-fit 3+1 hypothesis with the data from Fig.~\ref{fig:IC_rates}, to finally see how the best 
3+1 hypothesis compares to data.
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/final_rate_plot.pdf}
    \caption{The predicted binned event count assuming our best-fit 3+1 hypothesis in blue, compared with the 
    null 3+0 hypothesis in red and data points in black.}\label{fig:final_rate_plot}
\end{figure}

In Fig.~\ref{fig:final_rate_plot}, we see that the simulations are consistent with the 3+0 hypothesis with deviations of 6\% at high single-digit 
\si{\TeV} energies.

%TODO: Fix label sizes. 
%TODO: Include figure on theta34 and say that IC can’t distinguish it.

### 2021-12-11 thesis/tex/sterileTheory.tex
update

\subsection{Effect of Sterile Neutrino on Probabilities}
Since the new, sterile, neutrino does not interact weakly, how do we know if it's there? If the sterile mixing angle $\theta_{i4}$ is non-zero, 
we allow the sterile mass state to mix with the active state $i$. The most interesting case is when $\theta_{24} \neq 0$, 
which for $\dm[41] \sim \si{\eV \squared}$ gives rise to a resonant disappearance in the for $P_{\bar{\mu}\bar{\mu}}$, shown in Fig.~\ref{fig:sterile_resonance}.

@@

Keeping in mind that the event count from Eq.~\ref{eq:ICevents} scales as $E^{-1.7}$,
signals at higher energies will be more difficult to detect. Since higher values of $\dm[41]$ shifts the $\Pamam$ resonance peak to higher 
energies, high $\dm[41]$ values will be harder to rule out because the event count is already lower, as compared to energies 
at single-digit \si{\TeV} values. There is then a range of parameters, in which our detector will 
be able to distinguish between the two hypotheses. Using the two-neutrino approximation\footnote{In this approximation, we only consider mixing between $\nm$ and $\ns$, essentially a `1+1' picture.}, 
we can rewrite the expression for MSW resonance from Eq.~\ref{eq:MSW} as 
\begin{align}
    E^* = \frac{\dm[41] \cos{2\theta_{24}}}{2\sqrt{2} G_F N_e}\,,
\end{align}
which for $\dm[41] = \SI{1}{\TeV^2}$ and $\theta_{24} = 0.04$, gives us a resonance at $E^* = \SI{2.58}{\TeV}$, matching the $3+1$ simulation shown in
Fig.~\ref{fig:sterile_resonance}. From the two-neutrino picture, 
we see that the resonance peak in the $\anm$ channel scales linearly with $\dm[41]$, while the mixing angle dependence 
shows up as $\cos{2\theta_{24}}$, which is between 0.987 and 0.999 for the $\theta_[24}$ values considered in Fig.~\ref{fig:resonance_shift}.
Since our $\theta_{24}$ values considered are so low, the location of the resonance peak is essentially only dependent on the value of $\dm[41]$.
Thus, we expect a more stringent constraint of low values of $\dm[41]$, regardless of the value of $\theta_{24}$.

>>>>

\subsection{The Probability Effect of a Sterile Neutrino}
Since the new, sterile, neutrino does not interact weakly, how do we know if it's there? If the sterile mixing angle $\theta_{i4}$ is non-zero, 
we allow the sterile mass state to mix with the active state $i$. The most interesting case is when $\theta_{24} \neq 0$, 
which for $\dm[41] \sim \si{\eV \squared}$ gives rise to a resonant disappearance in the for $P_{\bar{\mu}\bar{\mu}}$, shown in Fig.~\ref{fig:sterile_resonance}.

@@

Keeping in mind that the event count from Eq.~\ref{eq:ICevents} scales as $E^{-1.7}$,
signals at higher energies will be more difficult to detect. Since higher values of $\dm[41]$ shifts the $\Pamam$ resonance peak to higher 
energies, high $\dm[41]$ values will be harder to rule out because the event count is already lower, as compared to energies 
at single-digit \si{\TeV} values. There is then a range of parameters, in which our detector will 
be able to distinguish between the two hypotheses. Using the two-neutrino approximation\footnote{In this approximation, we only consider mixing between $\nm$ and $\ns$, essentially a `1+1' picture.}, 
we can rewrite the expression for MSW resonance from Eq.~\ref{eq:MSW} as 
\begin{align}
    E^* = \frac{\dm[41] \cos{2\theta_{24}}}{2\sqrt{2} G_F N_e}\,,
\end{align}
which for $\dm[41] = \SI{1}{\TeV^2}$ and $\theta_{24} = 0.04$, gives us a resonance at $E^* = \SI{2.58}{\TeV}$, matching the $3+1$ simulation shown in
Fig.~\ref{fig:sterile_resonance}. From the two-neutrino picture, 
we see that the resonance peak in the $\anm$ channel scales linearly with $\dm[41]$, while the mixing angle dependence 
shows up as $\cos{2\theta_{24}}$, which is between 0.987 and 0.999 for the $\theta_[24]$ values considered in Fig.~\ref{fig:resonance_shift}.
Since our $\theta_{24}$ values considered are so low, the location of the resonance peak is essentially only dependent on the value of $\dm[41]$.
Thus, we expect a more stringent constraint of low values of $\dm[41]$, regardless of the value of $\theta_{24}$.
